---
title: "Charting in R"
author: ''
date: ''
output:
  ioslides_presentation:
    highlight: zenburn
    keep_md: no
    smaller: yes
    widescreen: yes
  beamer_presentation:
    highlight: tango
  slidy_presentation:
    highlight: tango
---

```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = FALSE)

```

## Aims and objectives

### Aim:

Expose you to the "grammar of graphics", to allow you create fancy charts in R

### Objectives:

- Introduce ggplot2
- Understand the grammar of constructing charts
- Practice creating plots yourself
- Signpost resources

## R Resources
- This workshop is based on the free eBook "R for Data Science" http://r4ds.had.co.nz (reproduced here under CC BY-NC-ND 3.0 US license)

- It covers all the basic usage of R and is pretty up to date.

## Part 1

The topics for this first part will include the following:

- Basic R Plotting
- Grammar of Graphics basics
- Aesthetics
- Facets
- Geoms

## Basic plots
- Base R has some basic plotting functionality, as illustrated here and on the following slide. However, it's quite limited in facility, and the current standard is to use the ggplot2 library, which provides a powerful grammar for creating charts.

Should you need them, the built in plot features can be called with

- Scatter: `plot()`
- Histogram: `hist()`
- Boxplot: `boxplot()`

----------

```{r}
# Basic Plot

# Set seed for plot
set.seed(5)

# Visualise plot
plot(rnorm(20), rnorm(20))
```

# ggplot2 - Basics

## Grammar of graphics

ggplot2 is a part of the Tidyverse, a popular collection of packages designed for data science.

Follows the "grammar of graphics" principles that plots can be composed from:

- A data set
- Mappings of variables to aesthetic features
- Geometric objects to represent the data ("geoms")
- Scales - properties of the visual mappings
- Facets (for small multiples)
- A coordinate system
- Annotations
- A theme (plot styling)


-----


We will use `package::function()` syntax throughout this course instead of `library()` calls.
This makes it easier to see where functions come from, avoids "function masking" and improves the
reproducibility and reuseability of code.

One exceptions to this is ggplot2's `aes()` function which is always nested inside other ggplot2 
functions. We will use the import package to load the `aes()` function as well as the `mpg` dataset:

```{r, highlight=TRUE}
import::from("ggplot2", "aes", "mpg")
```

We will also import the magrittr pipe:

```{r, highlight=TRUE}
import::from("magrittr", "%>%")
```

## Loading the data

For this intro we'll use the dataset `mpg`, which is included in the ggplot2 package. It contains 
information on cars fuel efficiency. We can use `dplyr::glimpse(mpg)` or `head(mpg)` or `mpg` to 
have a quick look at the data.   
```{r}
# Examine the mpg dataset
dplyr::glimpse(mpg)
```

`displ` contains the cars' engine size in litres\
`hwy` contains the cars fuel efficiency on the motorway in miles per gallon

## Minimal scatter plot

- Essential elements:
  - data
  - aesthetic mappings with aes()
  - geom function

- Add components with the `+` operator

```{r, fig.height=3, fig.width=4}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_point()
```

## Setting the aesthetics

- `aes()`can be inside `ggplot()` or the `geom` function
- setting aesthetics in a geom function allows you to add layers with different data or aesthetic mappings

```{r, fig.height=3, fig.width=4}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_point() +
  ggplot2::geom_point(aes(x = displ, y = cty), colour = "blue")
```

## General chart template

```{r eval = FALSE}
# chart template
ggplot(data = <DATA>, aes(<MAPPINGS>) )  + 
  <GEOM_FUNCTION> +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION> +
  <SCALE_FUNCTION> +
  <THEME FUNCTION>
```

Required:

- data
- aesthetic mappings
- a geometric function

The other functions are optional, with default values used if the are not specified.

## Exercises: Section 1

1. Run `ggplot2::ggplot(data = mpg)` what do you see?
2. What does the `drv` variable describe? Read the help for mpg data set to find out by running `?mpg`
3. Make a scatter plot of `hwy` vs. `cty`
4. Make a scatter plot of `class` vs. `drv`. Why isn't this plot useful?

### Extension

To fix the problem with exercise 4, make a scatter plot of `class` vs `cty` and colour the points by 
`drv` (more on this in the next section)

## Aesthetics

There are plenty of ways to map aesthetics in ggplot.

For example, we could the use the point colour to show the class of vehicle.
```{r, fig.height=3, fig.width=6}
# Aesthetics
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, colour = class)) +
  ggplot2::geom_point()
```

---

ggplot2 looks at the data type to set a default colour scale - categorical variables get discrete colours, continuous variables get a continous colour scale. These can be customised using `scale_` functions (covered later).

```{r, fig.height=4, fig.width=8}
# ggplot2 looks at the data type to set a default colour scale
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, colour = year)) +
  ggplot2::geom_point()
```

---

You can also use shapes to distinguish between categories, such as vehicle class.

But, what's wrong with this plot?
```{r, fig.height=4, fig.width=8}
# You can also use shapes to distinguish between categories
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, shape = class, colour = class)) +
  ggplot2::geom_point()
```

---

Some aesthetics are only suitable in certain situations. For example, using `shape` for `class` isn't appropriate as, by default, shape only supports 6 distinct categories (you can override this manually). Notice what has happened to "suv" here:
```{r, fig.height=4, fig.width=8}
# Some aesthetics are only suitable in certain situations
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, shape = class, colour = class)) +
  ggplot2::geom_point()
```

## Scale functions

Use `scale_` functions to set visual properties of the aesthetics, such as colours and axis scales.

```{r, fig.height=3.5, fig.retina=2}
# Use scale_ functions to set properties of the aesthetics
dplyr::filter(mpg, manufacturer %in% c('audi', 'volkswagen')) %>% 
  ggplot2::ggplot(aes(x = displ, y = hwy, colour = manufacturer)) +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_colour_manual(values = c("audi" = "#12436D", "volkswagen" = "#28A197")) +
  ggplot2::scale_x_continuous(limits = c(1, 5))
```

Hint: use `colours()` to get the list of built-in colours. Use hex values to specify any colour `'#28b7b7'`. 

## Setting an aesthetic to a fixed value

You can also set an aesthetic to a fixed value, rather than being mapped to a data variable.

```{r, fig.height=4, fig.width=8}
# Set an aesthetic to a fixed value
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_point(colour = "orange")
```

## Exercises: Section 2

### 1

- What's wrong with this code? Why aren't the points blue?

```{r, fig.height=3, fig.width=5}
# What's wrong with this code? Why aren't the points blue?
ggplot2::ggplot(data = mpg) +
  ggplot2::geom_point(aes(x = displ, y = hwy, colour = "blue"))
```

---

### 2 
- Map a continuous variable to `colour`, `size` and `shape`. How do these aesthetics behave 
differently for categorical vs. continuous variables? Hint: use `dplyr::glimpse(mpg)` to identify 
variable types.

### 3
- What happens if you map the same variable to multiple aesthetics?

### 4
- What happens if you map an aesthetic to something other than a variable name, like a Boolean statement `aes(colour = displ < 5)`

## Facets

Facets are a useful way to separate out variables into subplots.

There are two ways of doing this; `ggplot2::facet_wrap()` or `ggplot2::facet_grid()`. The main argument here is an R 'formula', where the faceted variable is preceded by a tilde ~ (e.g. ~class). To facet multiple variables, they can be separated by a tilde (e.g. drv ~ cyl)

```{r, fig.height=3}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~ class, nrow = 2)
```

---

You can lay out multiple facets in a grid:

```{r}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_point() +
  ggplot2::facet_grid(drv ~ class)
```


# Other geoms

## Geoms
There are tons of geoms available in ggplot - the best resource for choosing an appropriate geom is 
the cheat sheet. This can be found at https://rstudio.github.io/cheatsheets/data-visualization.pdf


## geom_smooth()

`ggplot2::geom_smooth()` takes data points and returns a regression model with confidence intervals. For example, if we just replace `ggplot2::geom_point()` with `ggplot2::geom_smooth()`, we get a loess curve.  

```{r, fig.height=4}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_smooth()
```

---

We can add the points as well, using the + operator; and have added a line break using `ggplot2::geom_line()` with a constant x value (this can be used for example to show a significant event in a time series). Note that the order of layers determines their order on the plot. As points are after the smooth line, they will be draw on top (and obscure) the line. 

```{r, fig.height=3}
# We can add the points as well, using the + operator
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_smooth() +
  ggplot2::geom_point() +
  ggplot2::geom_line(aes(x = 4.5))
```

## Exercise: Section 3

### 1

**Can you reproduce this plot?** 
Note that the points vary in `colour` based on the cars `class` but the smooth line does not.

```{r, fig.height=3, echo=FALSE}
# Note that the points vary in `colour` based on the cars `class` but the smooth line does not.
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_smooth() +
  ggplot2::geom_point(aes(colour = class))
```

---

### Extension

Try removing the confidence interval from the smooth line. Also try changing the linetype to dashed or dotted.

## Part 2

The topics for the second part will include:

- More geoms
- Chart Positioning 
- Themes, titles and Multiple Plots


## Bar charts
Counts of a single discrete variable.

```{r, fig.height=3, fig.width=9}
# Bar charts counting rows in the data
ggplot2::ggplot(data = mpg, aes(x = class)) +
  ggplot2::geom_bar()
```

---

Bar charts from tabulated data.

```{r, fig.height=3, fig.width=9}
# Bar charts from tabulated data
mpg %>% 
  dplyr::group_by(class) %>% 
  dplyr::summarise(count = dplyr::n()) %>% 
  ggplot2::ggplot(aes(x = class, y = count)) +
  ggplot2::geom_col() 
```


# Positioning

## Bar chart positioning

How could this chart be made easier to use?

Note: With bar charts, `fill =` determines the colour inside the bar and `colour =` controls the lines around the bar.

```{r, fig.height=3, fig.width=9}
# Bar chart positioning
ggplot2::ggplot(data = mpg, aes(x = manufacturer, fill = class)) +
  ggplot2::geom_bar(colour = 'black') +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0, hjust = 1))
```

---

By default bars are "stacked" on top of each other. This makes comparing proportions rather difficult. You can control this with the position argument. 'dodge' places bars next to each other, whilst 'fill' makes them all the same length (to allow for easier comparison).

```{r, fig.height=3}
# By default bars are "stacked" on top of each other. This makes comparing proportions rather difficult. 
# You can control this with the position argument.
ggplot2::ggplot(data = mpg, aes(x = manufacturer, fill = class)) +
  ggplot2::geom_bar(colour = 'black', position = 'dodge') +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0, hjust = 1))
```

## Density plots and histograms

Histograms are a great tool for summarizing and describing a continuous variable.

```{r, fig.height=3}
# Histogram
ggplot2::ggplot(data = mpg, aes(x = hwy)) +
  ggplot2::geom_histogram(bins = 30) 
```

---

Density plots are another nice way to visualise a continuous variable.

```{r, fig.height=3}
# Density plot
ggplot2::ggplot(data = mpg, aes(x = hwy)) +
  ggplot2::geom_density() 
```

---

We can use the `alpha` argument to reveal overlapping areas if we plot by group

```{r, fig.height=3}
# Overlapping density plots
ggplot2::ggplot(data = mpg, aes(x = hwy, fill = drv)) +
  ggplot2::geom_density(alpha = 0.5) 
```

## Boxplots and violin plots

```{r, fig.height=3}
# Box plots
ggplot2::ggplot(data = mpg, aes(x = drv, y = hwy)) +
  ggplot2::geom_boxplot(fill = "red")
```

---

```{r, fig.height=3}
# Violin plots
ggplot2::ggplot(data = mpg, aes(x = drv, y = hwy)) +
  ggplot2::geom_violin(aes(fill = drv))
```

## Line charts

US unemployment population data

```{r, fig.height=3}
# Line charts
ggplot2::economics %>% 
  ggplot2::ggplot(aes(x = date, y = unemploy)) +
  ggplot2::geom_line()
```


---

What is wrong with this plot?

```{r}
# What is wrong with this plot?
ggplot2::ggplot(data = mpg, aes(x = cty, y = hwy)) +
  ggplot2::geom_point()
```

---

Because there are multiple observations, some are **overplotted**. To correct this, you can add some random noise to the data with `position = "jitter"` or `ggplot2::geom_jitter()`. This is a bit of a compromise - you either have a chart that is accurate but suffers from over plotting, or one that contains some random noise but reveals the size of the data.

```{r, fig.height=3}
# Because there are multiple observations, some are **overplotted**. To correct this, you can add 
# some random noise to the data with `position="jitter"` or `geom_jitter`.
ggplot2::ggplot(data = mpg, aes(x = cty, y = hwy)) +
  ggplot2::geom_jitter(colour = 'red') +
  ggplot2::geom_point()
```

## Exercises: Section 4

1. Using the `mpg` dataset create a histogram of `cty`. What impact do different values for the `bins` argument have?

2. Make a density plot of `cty` by `drv`. Use the `alpha` argument to make the overlapping regions visible.

3. Recreate the US unemployed population line graph. Add another line to show the total population. 

*Hint* You will need to supply a new geom layer with a new aesthetic mapping.

# Themes, titles, and multiple plots

## Titles
Labels and titles can be added using the labs command.

```{r, fig.height=3}
# Themes, titles, and multiple plots
ggplot2::ggplot(data = mpg, aes(x = class, y = ..prop.., group = 1)) +
  ggplot2::geom_bar() +
  ggplot2::labs(title = "Proportion of sample by class", x = "Class", y = "Proportion")
```

## Multiple plots

Plots can be arranged using the `grid.arrange` command from the `gridExtra` package. First we store the plots in a variable using the `<-` operator.
```{r}
# Multiple plots
plot1<-
  ggplot2::ggplot(data = mpg, aes(x = cyl, y = ..prop.., group = 1)) +
  ggplot2::geom_bar(fill = "red") +
  ggplot2::labs(title = "Proportion of sample by cylinder", 
                x = "Cylinder", 
                y = "Proportion", 
                subtitle = " ")

plot2<-
  ggplot2::ggplot(data = mpg, aes(x = cty, y = hwy)) +
  ggplot2::geom_jitter(colour = 'red') +
  ggplot2::labs(title = "Highway fuel efficiency\number of cylinders", 
                subtitle = "Note, the points are jittered", 
                x = "Number of cylinders", 
                y = "Fuel efficiency")

```

---

```{r}
# gridExtra
gridExtra::grid.arrange(plot1, plot2, nrow = 1)

```


## Themes
You can change the theme to a number of presets.

---

```{r, fig.height=5, fig.width=8}
# Themes
gridExtra::grid.arrange(
  plot2 + ggplot2::theme_bw(), plot2 + ggplot2::theme_classic(), 
  plot2 + ggplot2::theme_dark(), plot2 + ggplot2::theme_light())
```

---

You can also make your own custom themes; plot are made up of four elements `element_text`, `element_line`, `element_rect`, and `element_blank`. Plots can be modified using these element commands. For example:

```{r, fig.height=3}
# You can also make your own custom themes
ugly_theme <-
  ggplot2::theme(
    text = ggplot2::element_text(colour = 'orange', face = 'bold'),
    panel.grid.major = ggplot2::element_line(colour = "violet", linetype = "dashed"),
    panel.grid.minor = ggplot2::element_blank(),
    panel.background = ggplot2::element_rect(fill = 'black', colour = 'red'),
  )

```

---

```{r, fig.height=3}
plot2 + ugly_theme
```

---

Pre-built MoJ and Government Analysis Function themes and colour schemes are available in the `mojchart` package.

```{r, fig.height=3}
# MoJ colour scheme
ggplot2::ggplot(data = mpg, aes(x = class, fill = drv)) +
  ggplot2::geom_bar(position = "dodge") +
  mojchart::scale_fill_moj(n = 3) +
  mojchart::theme_gss(xlabel = TRUE)
```

---

The Government Analysis Function colour scheme is accessible.
```{r, fig.height=3}
# Accessible colour scheme
ggplot2::ggplot(data = mpg, aes(x = class, fill = drv)) +
  ggplot2::geom_bar(position = "dodge") +
  mojchart::scale_fill_moj(n = 3, scheme = "govanal_bars") +
  mojchart::theme_gss(xlabel = TRUE)
```

## Exercises: Section 5

1. Create a basic scatter plot showing highway fuel efficiency vs city fuel efficiency. Add a theme_minimal; and colour the points based on the number of cylinders the car has.
    + *Hint*: here is some code to start you off:
    ```{r, eval=FALSE, results=FALSE}
    # Exercises: Section 5
    ggplot2::ggplot(data = mpg) +
       ggplot2::geom_point(mapping = aes(x = hwy, y = cty, colour = as.factor(cyl)))
    ```
2. Change the title and axis labels to something sensible based on what the chart is showing.
3. Create a custom theme changing:
    + major gridlines to grey, dashed
    + removing minor gridlines
    + x and y axis text to black, pt 10, italic
    + move the legend position to the bottom of the plot
4. Instead of the custom theme, use the gss theme from `mojchart` and one of the built-in colour schemes. 
    + (Number 4 may be homework if we are running out of time!)

## Saving plots

Most of the time plots will be created directly into an R Markdown output, or a shiny app. However plots can also be saved as image (png) file:

* 'Export' button in RStudio viewer

* `ggplot2::ggsave(filename = "plotname.png", plot = myplot)` - saves the plot into your current working directory in R Studio. Can then be downloaded from the platform via 'More' -> 'Export...'

In R Markdown or R Shiny: ggplot call will create the plot in situ.

## Resources

### References

- Cheatsheet: https://rstudio.github.io/cheatsheets/data-visualization.pdf
- ggplot2 website: https://ggplot2.tidyverse.org/
- ggplot2 book: https://ggplot2-book.org/

### Extras

- mojchart: https://github.com/moj-analytical-services/mojchart
- The R Graph Gallery: https://www.r-graph-gallery.com/index.html
- Further resources: https://trello.com/c/id9ir0Mf/18-r-charting-mapping-etc

### Alternatives

- Plotly: Plotly is a Javascript framework for creating interactive charts. The plotly R library 
has its own grammar for constructing charts, with power and flexibility rivaling that of ggplot2. 
However, it also has a function, ggplotly(), which can turn a ggplot chart into an interactive 
chart. Find out more here: https://plot.ly/r/


# Any questions? 




