---
title: "Data visualization in R: Solutions"
author: "Will Bowditch"
date: "10 February 2017"
output:
  ioslides_presentation:
    highlight: zenburn
    keep_md: no
    smaller: yes
    widescreen: yes
  slidy_presentation:
    highlight: tango
---

```{r, include=FALSE}
require(knitr)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, cache=FALSE)
require(tidyverse)
data(mpg)

```

## 1

```{r}
ggplot(data=mpg)
```

##2
```{r, echo=TRUE}
?mpg
```


##3 
Make a scatter plot of `hwy` vs. `cty`

```{r, fig.height=3}
ggplot(data=mpg, aes(x=hwy, y=cty))+
  geom_point()
```

##4
Make a scatter plot of `class` vs. `drv`. Why isn't this plot useful?

Note, `qplot` can be used for quick plotting with less customisation. 
```{r, fig.height=3}
qplot(mpg$class, mpg$drv, geom='point')
```

## 5 {}

 What's wrong with this code? Why aren't the points blue?
<div class="columns-2">
```{r, fig.height=3, fig.width=5}
ggplot(data = mpg)+
  geom_point(aes(x=displ, y=hwy, color = "blue"))

ggplot(data = mpg)+
  geom_point(aes(x=displ, y=hwy), color = "blue")


```
</div>


## 6 

Map a continuous variable to `color`, `size` and `shape`. How do these aesthetics behave differently for categorical vs. continuous variables? Hint: use `glimpse(mpg)` to identify variable types.


<div class="columns-2">
```{r, fig.height=3}
ggplot(data=mpg, aes(x=displ, y=hwy, color=cty))+
  geom_point()
```

```{r, fig.height=3}
ggplot(data=mpg, aes(x=displ, y=hwy, color=drv))+
  geom_point()
```

</div>

## 7

- What happens if you map the same variable to multiple aesthetics?

```{r,fig.height=3}
ggplot(data=mpg, aes(x=displ, y=hwy, color=drv, shape=drv))+
  geom_point()
```

## 8 

- What happens if you map an aesthetic to something other than a variable name, like a Boolean statement `aes(colour = displ <5)`

```{r, fig.height=3}
ggplot(data = mpg, aes(x= displ, y=hwy, color = displ <5))+
  geom_point()
```

## 9

You may have got this instead:
```{r, fig.height = 3}
ggplot(data = mpg, aes(x=displ, y=hwy, color=class))+
  geom_smooth()+
  geom_point()
```

---

Code to reproduce plot:
```{r, fig.height = 3}
ggplot(data = mpg, aes(x=displ, y=hwy))+
  geom_smooth()+
  geom_point(aes(color=class))
```



## 10

Produce a **scatter plot** between **displ** and **hwy** that uses `stat_smooth()` with a **general linear model** line of best fit.

```{r}
ggplot(data = mpg, aes(x=displ, y=hwy))+
  geom_point()+
  stat_smooth(method='glm')
```

