---
title: "R Charting: Solutions"
author: ""
date: ""
output:
  slidy_presentation:
    highlight: tango
  ioslides_presentation:
    highlight: zenburn
    keep_md: no
    smaller: yes
    widescreen: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, cache=FALSE)
import::from("ggplot2", "aes", "mpg")
import::from("magrittr", "%>%")

```

## Solutions: Section 1

### 1

```{r}
ggplot2::ggplot(data = mpg)
```

### 2
```{r, echo=TRUE}
?mpg
```


### 3 
Make a scatter plot of `hwy` vs. `cty`

```{r, fig.height=3}
ggplot2::ggplot(data = mpg, aes(x = hwy, y = cty)) +
  ggplot2::geom_point()
```

### 4
Make a scatter plot of `class` vs. `drv`. Why isn't this plot useful?

Note, `qplot` can be used for quick plotting with less customisation. 
```{r, fig.height=3}
ggplot2::qplot(mpg$class, mpg$drv, geom = "point")
```

### Extention
To fix the problem with exercise 4, make a scatter plot of `class` vs `cty` and colour the points by `drv` (more on this in the next section)

```{r}
ggplot2::ggplot(data = mpg) + 
  ggplot2::geom_point(mapping = aes(x = class, y = cty, colour = drv))
```

## Solutions: Section 2

### 1

What's wrong with this code? Why aren't the points blue?
<div class="columns-2">
```{r, fig.height=3, fig.width=5}
ggplot2::ggplot(data = mpg) +
  ggplot2::geom_point(aes(x = displ, y = hwy, colour = "blue"))

ggplot2::ggplot(data = mpg) +
  ggplot2::geom_point(aes(x = displ, y = hwy), colour = "blue")
```
</div>


### 2

Map a continuous variable to `colour`, `size` and `shape`. How do these aesthetics behave differently for categorical vs. continuous variables? Hint: use `dplyr::glimpse(mpg)` to identify variable types.


<div class="columns-2">

```{r, fig.height=3, fig.width=4}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, colour = cty)) +
  ggplot2::geom_point()
```


```{r, fig.height=3, fig.width=4}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, colour = drv)) +
  ggplot2::geom_point()
```

</div>

### 3

What happens if you map the same variable to multiple aesthetics?

```{r,fig.height=3}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, colour = drv, shape = drv)) +
  ggplot2::geom_point()
```

### 4

What happens if you map an aesthetic to something other than a variable name, like a Boolean statement `aes(colour = displ < 5)`

```{r, fig.height=3}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, colour = displ < 5)) +
  ggplot2::geom_point() 
```

#### 4.1

You may have got this instead:
```{r, fig.height = 3}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy, colour = class)) +
  ggplot2::geom_smooth() +
  ggplot2::geom_point()
```

## Solutions: Section 3


### Code to reproduce plot:
```{r, fig.height = 3}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_smooth() +
  ggplot2::geom_point(aes(colour = class))
```


###  Extention

Removing conf. intervals and changing  the line type

```{r, fig.height = 3}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_smooth(se = FALSE, linetype = "dotted", colour = "black") +
  ggplot2::geom_point(aes(colour = class))
```


## Solutions: Section 4 

### 1

Using the mpg dataset create a bar chart showing a count of the dataset grouped by class of car.

```{r, eval=FALSE}
ggplot2::ggplot(data = mpg) + 
  ggplot2::geom_bar(mapping = aes(x = class))
```

### 2 

Now specify a y co-ordinate mapping to show the average fuel displacement of each class of car.


```{r, fig.height = 2.7}
ggplot2::ggplot(data = mpg) + 
  ggplot2::geom_bar(mapping = aes(x = class, y = displ), stat = "summary", fun = "mean")
```

---

### 3

Add a fill to the bars to show the average split by the type of drive of each class of car. 

```{r, fig.height = 3}
ggplot2::ggplot(data = mpg) + 
  ggplot2::geom_bar(mapping = aes(x = class, y = displ, fill = drv), stat = "summary", 
                    fun = "mean", position = "dodge")
```

---

## Solutions: Section 5

### 1

Create a basic scatter plot showing highway fuel efficiency vs city fuel efficiency. Add a theme_minimal; and colour the points based on the number of cylinders the car has.

Change the title and axis labels to something sensible based on what the chart is showing.

```{r, fig.height = 3}

ggplot2::ggplot(data = mpg) +
  ggplot2::geom_point(mapping = aes(x = hwy, y = cty, colour = as.factor(cyl))) + 
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Fuel efficiency comparison between city and highway", 
                x = "Highway efficiency", y = "City efficiency")

```

### 2

Create a custom theme changing: major gridlines to grey, dashed; removing minor gridlines; x and y axis text to black, pt 10, italic; move the legend position to the bottom of the plot.

```{r, fig.height = 3}

ggplot2::ggplot(data = mpg) +
  ggplot2::geom_point(mapping = aes(x = hwy, y = cty, colour = as.factor(cyl))) + 
  ggplot2::labs(title = "Fuel efficiency comparison between city and highway", x = "Highway efficiency", y = "City efficiency") +
  ggplot2::theme(panel.grid.major = ggplot2::element_line(colour = "grey", linetype = "dashed"),
                 panel.grid.minor = ggplot2::element_blank(),
                 axis.text = ggplot2::element_text(colour = "black", size = 10, face = "italic"),
                 legend.position = "bottom")

```


## Appendix




### More on scatter plots and other geoms 

Produce a **scatter plot** between **displ** and **hwy** that uses `ggplot2::stat_smooth()` with a **general linear model** line of best fit.


```{r, fig.height=3}
ggplot2::ggplot(data = mpg, aes(x = displ, y = hwy)) +
  ggplot2::geom_point() +
  ggplot2::stat_smooth(method = 'glm')
```

---

### Stats and Geoms

Create a boxplot of city full efficiency by manufacturer. Expirement with changing the whisker length; change the colour of the outliers to red; and vary the width of the boxes proportional to the number of observations (hint: use ?stat_boxplot).

```{r}

ggplot2::ggplot(data = mpg, aes(x = manufacturer, y = cty)) +
  ggplot2::stat_boxplot(coef = 1, outlier.colour = "red", varwidth = TRUE)

```



### Extention: Section 4

Manually change the colours of the fill being used. Also change the font size and angle of the x axis text labels so that these can be read in the viewer.


```{r, fig.height = 3}

ggplot2::ggplot(data = mpg) + 
  ggplot2::geom_bar(mapping = aes(x = class, y = displ, fill = drv), stat = "summary", 
                    fun = "mean", position = "dodge") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, size = 8)) + 
  ggplot2::scale_fill_manual(values = c("blue", "darkblue", "blueviolet"))

```


### Scales and additional plot control

Try changing the x and y axis so that the scale starts at zero. 

Change the title of the legend to 'Number of Cylinders' and re-name the labels with the text version of the number (e.g. 'four' instead of '4')

```{r, fig.height = 3}

ggplot2::ggplot(data = mpg) +
  ggplot2::geom_point(mapping = aes(x = hwy, y = cty, colour = as.factor(cyl))) + 
  ggplot2::labs(title = "Fuel efficiency comparison between city and highway", 
                x = "Highway efficiency", 
                y = "City efficiency") +
  ggplot2::scale_x_continuous(expand = c(0,0)) +
  ggplot2::scale_y_continuous(expand = c(0,0)) +
  ggplot2::expand_limits(x = c(0, 50), y = c(0, 40)) +
  ggplot2::scale_colour_discrete(labels = c("Four", "Five", "Six", "Eight"), name = "Number of Cylinders")

```

### Linetypes, colours etc

#### Putting it all together

Plot a line chart to show change in city efficiency between the two years, with a different line colour and line type for each class. Make sure you specify the dataset as mpg_group. 

```{r, eval = T}
# mpg_group <- dplyr::select(mpg, class, year, cty) %>% 
#   dplyr::group_by(class, year) %>% 
#   dplyr::summarise(mean = mean(.data$cty))
# 
# ggplot2::ggplot(data = mpg_group) + 
#   ggplot2::geom_line(mapping = aes(x = year, y = mean, colour = class, linetype = class)) +
#   ggplot2::scale_x_continuous(breaks = c(1999, 2008), labels = c("1999", "2008")) +
#   ggplot2::scale_y_continuous(limits = c(10,22.5), breaks = seq(10, 22.5, 2.5))

```

#### Introducing breaks in axis

```{r, fig.height = 6, echo = TRUE}
# 
# ggplot2::ggplot(data = mpg_group) + 
#   ggplot2::geom_line(mapping = aes(x = year, y = cty, colour = class, linetype = class)) +
#   ggplot2::scale_x_continuous(breaks = c(1999, 2008), labels = c("1999", "2008")) +
#   ggplot2::scale_y_continuous(limits = c(10,22.5), breaks = seq(10, 22.5, 2.5))
```


#### Changing line type

Experiment with changing the line types manually, and changing the thickness of the lines to make them larger.

```{r, fig.height = 4, eval=TRUE}
# 
# ggplot2::ggplot(data = mpg_group) +
#   ggplot2::geom_line(mapping = aes(x = year, y = cty, colour = class, linetype = class), size = 1.25) +
#   ggplot2::scale_x_continuous(breaks = c(1999, 2008), labels = c("1999", "2008")) +
#   ggplot2::scale_linetype_manual(values = c(4,1,6,2,3,5,9)) +
#   ggplot2::scale_y_continuous(limits = c(10,22.5), breaks = seq(10, 22.5, 2.5))

```






