---
title: "R Charting: Solutions"
author: ""
date: ""
output:
  ioslides_presentation:
    highlight: zenburn
    keep_md: no
    smaller: yes
    widescreen: yes
  slidy_presentation:
    highlight: tango
---

```{r, include=FALSE}
require(knitr)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, cache=FALSE)
require(tidyverse)
require(ggthemes)
data(mpg)

```

## 1

```{r}
ggplot(data=mpg)
```

##2
```{r, echo=TRUE}
?mpg
```


##3 
Make a scatter plot of `hwy` vs. `cty`

```{r, fig.height=3}
ggplot(data=mpg, aes(x=hwy, y=cty))+
  geom_point()
```

##4
Make a scatter plot of `class` vs. `drv`. Why isn't this plot useful?

Note, `qplot` can be used for quick plotting with less customisation. 
```{r, fig.height=3}
qplot(mpg$class, mpg$drv, geom='point')
```

---

To fix the problem with exercise 4, make a scatter plot of `class` vs `cty` and colour the points by `drv` (more on this in the next section)

```{r}

ggplot(data = mpg) + geom_point(mapping = aes(x = class, y = cty, colour = drv))

```


## 5

What's wrong with this code? Why aren't the points blue?
<div class="columns-2">
```{r, fig.height=3, fig.width=5}
ggplot(data = mpg)+
  geom_point(aes(x=displ, y=hwy, color = "blue"))

ggplot(data = mpg)+
  geom_point(aes(x=displ, y=hwy), color = "blue")


```
</div>


## 6 

Map a continuous variable to `color`, `size` and `shape`. How do these aesthetics behave differently for categorical vs. continuous variables? Hint: use `glimpse(mpg)` to identify variable types.


<div class="columns-2">

```{r, fig.height=3, fig.width=4}
ggplot(data=mpg, aes(x=displ, y=hwy, color=cty))+
  geom_point()
```


```{r, fig.height=3, fig.width=4}
ggplot(data=mpg, aes(x=displ, y=hwy, color=drv))+
  geom_point()
```

</div>

## 7

What happens if you map the same variable to multiple aesthetics?

```{r,fig.height=3}
ggplot(data=mpg, aes(x=displ, y=hwy, color=drv, shape=drv))+
  geom_point()
```

## 8 

What happens if you map an aesthetic to something other than a variable name, like a Boolean statement `aes(colour = displ <5)`

```{r, fig.height=3}
ggplot(data = mpg, aes(x= displ, y=hwy, color = displ <5))+
  geom_point()
```

## 9

You may have got this instead:
```{r, fig.height = 3}
ggplot(data = mpg, aes(x=displ, y=hwy, color=class))+
  geom_smooth()+
  geom_point()
```

---

Code to reproduce plot:
```{r, fig.height = 3}
ggplot(data = mpg, aes(x=displ, y=hwy))+
  geom_smooth()+
  geom_point(aes(color=class))
```

---

Extension:
```{r, fig.height = 3}
ggplot(data = mpg, aes(x=displ, y=hwy))+
  geom_smooth(se = FALSE, linetype = "dotted", colour = "black")+
  geom_point(aes(color=class))
```



## 10

Produce a **scatter plot** between **displ** and **hwy** that uses `stat_smooth()` with a **general linear model** line of best fit.


```{r, fig.height=3}
ggplot(data = mpg, aes(x=displ, y=hwy))+
  geom_point()+
  stat_smooth(method='glm')
```

---

Extension: Create a boxplot of city full efficiency by manufacturer. Expirement with changing the whisker length; change the colour of the outliers to red; and vary the width of the boxes proportional to the number of observations (hint: use ?stat_boxplot).

```{r}

ggplot(data=mpg, aes( x=manufacturer, y=cty)) +
   stat_boxplot(coef = 1, outlier.colour = "red", varwidth = TRUE)

```


##11

Using the mpg dataset create a bar chart showing a count of the dataset grouped by class of car.

```{r, eval=FALSE}

ggplot(data = mpg) + geom_bar(mapping = aes(x = class))

```

Now specify a y co-ordinate mapping to show the average fuel displacement of each class of car.

  
```{r, fig.height = 2.7}

ggplot(data = mpg) + geom_bar(mapping = aes(x = class, y = displ), stat = "summary", fun.y = "mean")

```

---

Add a fill to the bars to show the average split by the type of drive of each class of car. 

  
```{r, fig.height = 3}

ggplot(data = mpg) + geom_bar(mapping = aes(x = class, y = displ, fill = drv), stat = "summary", 
                              fun.y = "mean", position = "dodge")

```

---

Extension: Manually change the colours of the fill being used. Also change the font size and angle of the x axis text labels so that these can be read in the viewer.


```{r, fig.height = 3}

ggplot(data = mpg) + 
  geom_bar(mapping = aes(x = class, y = displ, fill = drv), stat = "summary", 
           fun.y = "mean", position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, size = 8)) + 
  scale_fill_manual(values = c("blue", "darkblue", "blueviolet"))

```

## 12

Create a basic scatter plot showing highway fuel efficiency vs city fuel efficiency. Add a google docs theme from the ggthemes package; and colour the points based on the number of cylinders the car has.

Change the title and axis labels to something sensible based on what the chart is showing.

```{r, fig.height = 3}

ggplot(data = mpg) +
  geom_point(mapping = aes(x = hwy, y = cty, colour = as.factor(cyl))) + 
  theme_gdocs() +
  labs(title = "Fuel efficiency comparison between city and highway", 
       x = "Highway efficiency", y = "City efficiency")

```

---

Create a custom theme changing: major gridlines to grey, dashed; removing minor gridlines; x and y axis text to black, pt 10, italic; move the legend position to the bottom of the plot.


```{r, fig.height = 3}

ggplot(data = mpg) +
  geom_point(mapping = aes(x = hwy, y = cty, colour = as.factor(cyl))) + 
  labs(title = "Fuel efficiency comparison between city and highway", 
       x = "Highway efficiency", 
       y = "City efficiency") +
  theme(panel.grid.major = element_line(colour = "grey", linetype = "dashed"),
        panel.grid.minor = element_blank(),
        axis.text = element_text(colour = "black", size = 10, face = "italic"),
        legend.position = "bottom")

```

---

Try changing the x and y axis so that the scale starts at zero. 

Change the title of the legend to 'Number of Cylinders' and re-name the labels with the text version of the number (e.g. 'four' instead of '4')

```{r, fig.height = 3}

ggplot(data = mpg) +
  geom_point(mapping = aes(x = hwy, y = cty, colour = as.factor(cyl))) + 
  labs(title = "Fuel efficiency comparison between city and highway", 
       x = "Highway efficiency", 
       y = "City efficiency") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  expand_limits(x = c(0, 50), y = c(0, 40)) +
  scale_colour_discrete(labels = c("Four", "Five", "Six", "Eight"), name = "Number of Cylinders")

```

## 13

Plot a line chart to show change in city efficiency between the two years, with a different line colour and line type for each class. Make sure you specify the dataset as mpg_group. 

```{r, eval = FALSE}

library(dplyr)
mpg_group    <- mpg %>% select(class, year, cty) %>% group_by(class, year) %>% summarise_all(mean)

ggplot(data = mpg_group) + 
  geom_line(mapping = aes(x = year, y = cty, colour = class, linetype = class)) +
  scale_x_continuous(breaks = c(1999, 2008), labels = c("1999", "2008")) +
  scale_y_continuous(limits = c(10,22.5), breaks = seq(10, 22.5, 2.5))

```

---

```{r, fig.height = 6, echo = FALSE}

mpg_group    <- mpg %>% select(class, year, cty) %>% group_by(class, year) %>% summarise_all(mean)

ggplot(data = mpg_group) + 
  geom_line(mapping = aes(x = year, y = cty, colour = class, linetype = class)) +
  scale_x_continuous(breaks = c(1999, 2008), labels = c("1999", "2008")) +
  scale_y_continuous(limits = c(10,22.5), breaks = seq(10, 22.5, 2.5))
```


---

Experiment with changing the line types manually, and changing the thickness of the lines to make them larger.

```{r, fig.height = 4}

ggplot(data = mpg_group) + 
  geom_line(mapping = aes(x = year, y = cty, colour = class, linetype = class), size = 1.25) +
  scale_x_continuous(breaks = c(1999, 2008), labels = c("1999", "2008")) +
  scale_linetype_manual(values = c(4,1,6,2,3,5,9)) +
  scale_y_continuous(limits = c(10,22.5), breaks = seq(10, 22.5, 2.5))

```

